\chapter{Introduction}
 
Since the C:DDA-Discord grows each day, the community constantly raises beginners' questions regarding the start of a new game so I've decided to put out a (hopefully) comprehensive manual to Cataclysm: Dark Days Ahead, that will cover most of the things encountered in the game. This Text will, over the course of the creation, cover all the basic things: Character creation, The base game and it's most important controls, guides to the various stages of the game (played by me on basic settings) as well as several parts of the games' explanation in-depth.

The reasoning for me writing a guide, or at this point, tutorial rather to begin with is simple - all the information found for this game is incredibly dated and therefore I wish to put up a comprehensive knowledge-book for newcomers to be pointed to if they wish to learn more about the game but also for veterans to check out again if they lack some information or just want to refresh their memory after being absent for ages.

So before we start, some info about me: I'm Pr0man, a somewhat dated veteran of the game who played this back when it was just called "Cataclysm" (C:DDA is technically a fork of the original). I'm more often than not streaming the game on twitch and I usually play with some pretty hard world settings, so be aware that this guide may or may not be biased towards a more careful playstyle. I am certain that I have a good understanding of the general mechanics of the game, as well as some in-depth knowledge of some other intricacies the game can throw at you without warning.

HOWEVER - I am by no means all-knowing and therefore you should be careful that some topics may not contain correct or proper information to boot. I tried collecting knowledge from many different sources in order to compile them into this wordy document, but remember to take in all of that with a grain of salt. Feel free to correct me, to ask me certain things or whatever you feel could be done to improve this piece of work I try to call a "tutorial".

While the game has changed drastically over the course of its development, many questions have not. And after answering many questions over and over again, I decided it is time to put my wasted energy into writing this. And without further ado, let's get to the meat of things.